---
title: "EDS-223 Homework Assignment #2"
author: "Isabella Segarra"
date: 10-06-2025
format: 
  pdf: default 
  html: default
theme: pandoc
execute: 
  warning: false
  message: false
---

# Exploring Patterns of Environmental Justice

### Objective

There are three datasets used in this project: 
1. EJSCREEN: data is from the United States Environmental Protection Agency’s former EJScreen: Environmental Justice Screening and Mapping Tool. 
2. HOLC Redlining: data is from Digital Scholarship Lab at the University of Richmond depicting the Home Owners’ Loan Corporation (HOLC) New Deal policy and subsequent, redlining. 
3. Biodiversity observations: data is from the Global Biodiversity Information Facility and contains observations on bird abundance.

## Environment Set-up 
```{r}
#|output: false

# Load relevant libraries
library(sf) # For vector data
library(stars) # For raster data 
library(tmap) # For static and interactive maps
library(here) # For importing data 
library(tidyverse) # For data cleaning
library(dplyr) # For filtering data 


# If the libraries are not installed, please install the below packages: 

# install.packages("sf")
# install.packages("stars")
# install.packages("tmap")
# install.packages("here")
# install.packages("tidyverse")
# install.packages("dplyr")

```

## Load in Data 
```{r}
# Load in EJSCREEN data for data on census blocks 
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

# Load in HOLC Redlining data
redlining <- sf::st_read(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))

# Load in Bird Observations data 
bird_obs <- sf::st_read(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))

```

## Part 1: Legacy of redlining in current environmental (in)justice

### Objective: 
Explore historical redlining in Los Angeles and its legacy on present-day environmental justice.

```{r}
#......Create a map of historical redlining neighborhoods......

tm_shape(redlining) + # Map layer of redlined neighborhoods 
  tm_borders() + 
  tm_polygons(fill = "fill")+
  tm_basemap("Open Street Map")

```

